---
title: "Small Worlds and Large Worlds"
output: html_notebook
---

## 2.1 The garden of forking data

Bayesian inference is really just counting and comparing possibilities

  - Need to consider everything that could have happened (all forking paths)
  - Alternative sequences of events are cultivated and then compared to what happened
    - events that don't match are pruned
  
### 2.1.1 Counting possibilities

Toy example:

Picking blue and white marbles from a bag. In total there are four marbels but we
don't know how many are of each color. Five total possibilities:

  - 4w0b
  - 3w1b
  - 2w2b
  - 1w3b
  - 0w4b
 
We can call these possibilities *conjectures*. Goal is to find the conjecture that
is most plausible given some evidence - in this case a sequence of 3 marbles that
were pulled out of the bag.

Say we pick three marbles with replacement and observe **b-w-b** in this order.
We can visualize the possibilities branching outward as below:


```{r}
library(DiagrammeR)

# Create branching possibilities diagram for first pick
n = 5
node_df <- create_node_df(
  n = n,
  label="",
  style = c('invisible', rep('filled', n-1)),
  color = 'black',
  fillcolor = c(NA, 'blue', 'white', 'white', 'white'),
  shape = 'circle',
)

edge_df <- create_edge_df(
  from = c(1, 1, 1, 1),
  to = c(2, 3, 4, 5),
  arrowhead='none',
  color='black'
)

graph <- create_graph(
  nodes_df = node_df,
  edges_df = edge_df,
  attr_theme='bt'
) 

graph |> DiagrammeR::render_graph()

```

```{r}
library(DiagrammeR)

# Create branching possibilities diagram for second pick
n = 21
node_df <- create_node_df(
  n = n,
  label="",
  style = c('invisible', rep('filled', n-1)),
  color = 'black',
  fillcolor = c('black', rep(c('blue', 'white', 'white', 'white'), 5)),
  shape = 'circle',
)

edge_df <- create_edge_df(
  from = c(rep(1, 4), rep(2, 4), rep(3, 4),rep(4,4), rep(5,4)),
  to = 2:n,
  arrowhead='none',
  color='black'
)

graph <- create_graph(
  nodes_df = node_df,
  edges_df = edge_df,
  attr_theme='bt'
) 

graph |> render_graph()
```

```{r}
# Create branching possibilities diagram for all three pickes
n = 64+20+1
node_df <- create_node_df(
  n = n,
  label="",
  style = c('invisible', rep('filled', n-1)),
  color = 'black',
  fillcolor = c('black', rep(c('blue', 'white', 'white', 'white'), (n-1)/4)),
  shape = 'circle',
  fontcolor='black'
)

edge_df <- create_edge_df(
  from = rep(1:((n-1)/4), each=4),
  to = 2:n,
  arrowhead='none',
  color='black'
)

graph <- create_graph(
  nodes_df = node_df,
  edges_df = edge_df,
  attr_theme='bt'
) 

graph |> render_graph()
```

In all there are 4^3 = 64 different branches or possible paths. If we follow the paths there are
3 paths that can give us a draw of **b-w-b** assuming the bag has 3w1b.

```{r}
n = 64+20+1
node_df <- create_node_df(
  n = n,
  label="",
  style = c('invisible', rep('filled', n-1)),
  color = 'black',
  fillcolor = c('black', rep(c('blue', 'white', 'white', 'white'), (n-1)/4)),
  # penwidth = c(2, 1, 1, 1, 1, 2, 2, 2, rep(1, 16), 2, 1, 1, 1, 2, 1, 1, 1, 
  #              2,rep(1, 12*4 + 3)),
  shape = 'circle',
  fontcolor='black'
)

edge_df <- create_edge_df(
  from = rep(1:((n-1)/4), each=4),
  to = 2:n,
  arrowhead='none',
  color=c('black', rep('grey', 4), rep('black', 3), rep('grey', 16),
          rep(c('black', rep('grey', 3)), 3), rep('grey', 12*4)),
  penwidth = c(10, rep(1, 4), rep(10, 3), rep(1, 16),
          rep(c(10, rep(1, 3)), 3), rep(1, 12*4))
)

graph <- create_graph(
  nodes_df = node_df,
  edges_df = edge_df,
  attr_theme='bt'
) 

graph |> render_graph()
```

Inferential power comes from comparing this count (3/64) to the number of ways each of the other
conjectures of the bag's contents could produce the same data. We can do the same for each conjecture and get the following counts

```{r, layout="l-body-outset"}
library(knitr)

df <- data.frame(
  Conjecture = c("[W-W-W-W]",
                   "[B-W-W-W]",
                   "[B-B-W-W]",
                   "[B-B-B-W]",
                   "[B-B-B-B]"),
  `Ways to produce B-W-B` = c("0 x 4 x 0 = 0",
                              "1 x 3 x 1 = 3",
                              "2 x 2 x 2 = 8",
                              "3 x 1 x 3 = 9",
                              "4 x 0 x 4 = 0")
)

df
```

### 2.1.2 Combining other information

When new data becomes available - updated counts can be calculated by multiplying the prior counts with the number of ways each conjecture could produce the new observation.

  - e.g. say we pick another marble and it is blue, the new count for the conjecture 1w3b is (3x1x3)x3=27 
  - still works if new data is not the same type of data
    - Say factory lets you know that the ratio for the conjectures 3b1w:2b2w:1b3w is 1:2:3 then you could just multiply the prior count by this new information
    
```{r}
df <- data.frame(
  Conjecture = c("[W-W-W-W]",
                   "[B-W-W-W]",
                   "[B-B-W-W]",
                   "[B-B-B-W]",
                   "[B-B-B-B]"),
  `Ways to produce B` = c(0, 1, 2, 3, 4),
  `Prior counts` = c("0 x 4 x 0 = 0",
                              "1 x 3 x 1 = 3",
                              "2 x 2 x 2 = 8",
                              "3 x 1 x 3 = 9",
                              "4 x 0 x 4 = 0"),
  `New count` = c('0 x 0 = 0',
                  '3 x 1 = 3',
                  '8 x 2 = 16',
                  '9 x 3 = 27',
                  '0 x 4 = 0')
)

df
```


```{r}
df <- data.frame(
  Conjecture = c("[W-W-W-W]",
                   "[B-W-W-W]",
                   "[B-B-W-W]",
                   "[B-B-B-W]",
                   "[B-B-B-B]"),
  Ways_to_produce_B = c(0, 1, 2, 3, 4),
  Prior_count = c("0 x 4 x 0 x 0 = 0",
                              "1 x 3 x 1 x 1 = 3",
                              "2 x 2 x 2 x 2 = 16",
                              "3 x 1 x 3 x 3 = 27",
                              "4 x 0 x 4 x 4 = 0"),
  Factory_count = c(0, 3, 2, 1, 0),
  New_count = c('0 x 0 = 0',
                  '3 x 3 = 9',
                  '16 x 2 = 32',
                  '27 x 1 = 27',
                  '0 x 4 = 0')
)

df
```

### 2.1.3 From counts to probability

Mathematical way to formulate garden of forking data:

  - If $p$ is the proportion of marbles that are blue (so that 1b3w = 0.25)
  - and $D_{new}$ is the new data (e.g. B-W-B)
  - Then:
  
$$\text{plausibility of } p \text{ after }D_{new} \propto \text{ways } p \text{ can produce } D_{new} * \text{prior plausibility of } p$$

You can then convert these counts to probabilities by dividing by the sum of products

$$\text{plausibility of } p \text{ after }D_{new} = \frac{\text{ways } p \text{ can produce } D_{new} * \text{prior plausibility of } p}{\text{sum of products}}$$
For example:

```{r}
# Make Table
df <- data.frame(
  Conjecture = c("[W-W-W-W]",
                   "[B-W-W-W]",
                   "[B-B-W-W]",
                   "[B-B-B-W]",
                   "[B-B-B-B]"),
  p = c(0, 0.25, 0.5, 0.75, 1),
  Ways_to_produce_data = c(0, 3, 8, 9, 0),
  Plausibility = c(0, 0.15, 0.40, 0.45, 0)

)

# Calculate probability

ways = c(0, 3, 8, 9, 0)
print(ways/sum(ways))

df
```

Each of these peices has a direct partner in applied probability theory:

  - A conjectured proportion, $p$, is usually called a **PARAMETER** value
  - The relative number of ways that a value $p$ can produce the data is usually called a **LIKELIHOOD**
  - The prior plausibility of any specific $p$ is usually called the **PRIOR PROBABILITY**
  - The new, updated plausibility of any specific $p$ is usually called the **POSTERIOR PROBABILITY**